---
output: 
  github_document:
    html_preview: true
encoding: UTF-8
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
devtools::load_all()
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300
)
# Ensure UTF-8 encoding for proper Unicode character display
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
```

# brpubVJCE 📊

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/BR-Visualization/brpubVJCE/graph/badge.svg)](https://app.codecov.io/gh/BR-Visualization/brpubVJCE)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of brpubVJCE is to generate benefit-risk visualizations for the publication “How to visually integrate value judgment with clinical evidence”.

## Installation

### Recommended Installation

```r
# Install using pak (recommended)
install.packages("pak")
pak::pak("BR-Visualization/brpubVJCE")
```

### Alternative Installation  

```r
# Install using remotes
install.packages("remotes")
remotes::install_github("BR-Visualization/brpubVJCE")
```


## Quick Start

Here's how to create your first benefit-risk visualization:

```{r example, fig.width=10, fig.height=6}
create_forest_dot_plot(
  prepare_forest_dot_data(effects_table)
)
```


## Getting Help

- 📖 **Documentation**: Use `?create_forest_dot_plot` or `?prepare_forest_dot_data` for detailed function help
- 🐛 **Issues**: Report bugs at [GitHub Issues](https://github.com/BR-Visualization/brpubVJCE/issues)  
- 💬 **Discussions**: Join discussions at [GitHub Discussions](https://github.com/BR-Visualization/brpubVJCE/discussions)
- 📧 **Contact**: Reach out to the package maintainers via GitHub

## Citation

If you use this package in your research, please cite:

```{r citation, eval=FALSE}
citation("brpubVJCE")
```

## Contributing

We welcome contributions! Please see our [Contributing Guidelines](https://github.com/BR-Visualization/brpubVJCE/blob/main/CONTRIBUTING.md) for details on:

- 🔧 How to submit bug reports and feature requests
- 📝 How to contribute code and documentation  
- 🧪 How to run tests and ensure code quality
- 📋 Our code of conduct and style guidelines

## License

This package is licensed under the MIT License. See the [LICENSE](LICENSE.md) file for details.

---

*Built with ❤️ for the benefit-risk visualization community*
