---
title: "MCDA Benefit-Risk Analysis: Comprehensive Guide"
author: "BR Visualization Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```

# 1. Introduction

## 1.1 What is MCDA?

Multi-Criteria Decision Analysis (MCDA) is a quantitative method for comparing treatment options by:

- **Combining multiple outcomes** with different measurement scales
- **Assigning weights** to reflect relative importance of each outcome
- **Integrating clinical evidence with value judgment** transparently
- **Producing comparable scores** across treatment options

## 1.2 Why Use MCDA?

Traditional benefit-risk assessments compare outcomes individually, but real-world decisions require:

- Simultaneous consideration of multiple benefits and risks
- Recognition that some outcomes matter more than others
- Comparison across different measurement scales

## 1.3 Key Visualizations

This guide covers two complementary visualizations:

1. **Figure 7**: Step-by-step MCDA calculation (single drug example)
2. **Figure 8**: Stacked bar chart showing weighted contributions across all drugs
3. **Figure 9**: Benefit-risk map positioning treatments in 2D space

---

# 2. Setup and Dependencies

```{r libraries}
# library(MCDA)
library(ggplot2)
library(dplyr)
library(reshape2)
library(gridExtra)
library(knitr)
```

---

# 3. Data Preparation

## 3.1 Create Raw Performance Data

```{r create-data}
create_publication_data <- function() {
  # Treatment effects for each outcome (based on Figure 6)
  performance_data <- data.frame(
    Treatment = c("Placebo", "Drug A", "Drug B", "Drug C", "Drug D"),
    
    # Benefits (higher is better)
    Primary_Efficacy = c(0.25, 0.62, 0.45, 0.58, 0.52),
    Secondary_Efficacy = c(5, 42, 28, 38, 35),
    Quality_of_Life = c(2, 38, 25, 32, 42),
    
    # Risks (lower is better)
    Recurring_AE = c(0.15, 0.35, 0.25, 0.42, 0.20),
    Rare_SAE = c(0.02, 0.04, 0.03, 0.06, 0.02)
  )
  
  return(performance_data)
}

pub_data <- create_publication_data()
kable(pub_data, caption = "Table 1: Raw Treatment Performance Data", digits = 3)
```

**Purpose**: Establish baseline data including placebo and active treatments.

## 3.2 Calculate Treatment Differences vs Placebo

```{r calc-differences}
calculate_treatment_differences <- function(data) {
  placebo_row <- data[data$Treatment == "Placebo", ]
  
  differences <- data.frame(
    Treatment = data$Treatment[data$Treatment != "Placebo"],
    
    # Benefits: active - placebo (positive = improvement)
    Primary_Efficacy_Diff = data$Primary_Efficacy[data$Treatment != "Placebo"] - 
      placebo_row$Primary_Efficacy,
    Secondary_Efficacy_Diff = data$Secondary_Efficacy[data$Treatment != "Placebo"] - 
      placebo_row$Secondary_Efficacy,
    Quality_of_Life_Diff = data$Quality_of_Life[data$Treatment != "Placebo"] - 
      placebo_row$Quality_of_Life,
    
    # Risks: placebo - active (positive = less risk than placebo)
    Recurring_AE_Diff = placebo_row$Recurring_AE - 
      data$Recurring_AE[data$Treatment != "Placebo"],
    Rare_SAE_Diff = placebo_row$Rare_SAE - 
      data$Rare_SAE[data$Treatment != "Placebo"]
  )
  
  return(differences)
}

treatment_diffs <- calculate_treatment_differences(pub_data)
kable(treatment_diffs, caption = "Table 2: Treatment Differences vs Placebo", digits = 3)
```

**Why Treatment Differences?**

- MCDA requires all criteria oriented in same direction ("higher is better")
- For **benefits**: Drug - Placebo (positive = improvement over placebo)
- For **risks**: Placebo - Drug (positive = fewer AEs than placebo = good)
- Represents clinically meaningful comparisons: "How much better/worse is the drug vs placebo?"

## 3.3 Create Performance Matrix

```{r performance-matrix}
# Convert treatment differences to matrix format for MCDA
performance_matrix_pub <- as.matrix(treatment_diffs[, -1])
rownames(performance_matrix_pub) <- treatment_diffs$Treatment

# Clean column names
colnames(performance_matrix_pub) <- c("Primary_Efficacy", "Secondary_Efficacy",
                                       "Quality_of_Life", "Recurring_AE", "Rare_SAE")

kable(performance_matrix_pub, 
      caption = "Table 3: Performance Matrix for MCDA", 
      digits = 3)
```

**Key Point**: This matrix contains treatment differences, not absolute values. All values are now oriented "higher is better."

---

# 4. MCDA Configuration

## 4.1 Define Criterion Weights

```{r weights}
weights_pub <- c(
  Primary_Efficacy = 0.229,    # 22.9% - most important benefit
  Secondary_Efficacy = 0.057,  # 5.7%  - less important (correlated)
  Quality_of_Life = 0.115,     # 11.5% - patient outcome
  Recurring_AE = 0.479,        # 47.9% - MOST important (safety dominant)
  Rare_SAE = 0.120             # 12.0% - serious but rare
)

weights_df <- data.frame(
  Criterion = names(weights_pub),
  Weight = weights_pub,
  Percentage = paste0(round(weights_pub * 100, 1), "%"),
  Category = c("Benefit", "Benefit", "Benefit", "Risk", "Risk")
)
kable(weights_df, caption = "Table 4: Criterion Weights", row.names = FALSE)
```

**Weight Rationale**:

- **Recurring AE (47.9%)**: Safety paramount; affects many patients
- **Primary Efficacy (22.9%)**: Core outcome but balanced with safety
- **Rare SAE (12.0%)**: Serious but rare, moderate weight
- **Quality of Life (11.5%)**: Important patient-reported outcome
- **Secondary Efficacy (5.7%)**: Low to avoid double-counting with primary

**Note**: Weights should be elicited from stakeholders using structured methods (e.g., swing-weighting).

## 4.2 Normalize Performance Matrix

```{r normalize}
normalized_pub <- apply(performance_matrix_pub, 2, function(x) {
  (x - min(x)) / (max(x) - min(x))
})

kable(normalized_pub, 
      caption = "Table 5: Normalized Performance Matrix (0-1 scale)", 
      digits = 3)
```

**Why Normalize?**

- Outcomes have different scales (0.37 vs 40 for treatment differences)
- Normalization creates 0-1 scale where:
  - **0** = worst treatment difference (least improvement over placebo)
  - **1** = best treatment difference (most improvement over placebo)
- Makes outcomes comparable for weighted aggregation

**Example**: Drug A Primary Efficacy
- Raw difference: 0.37 (37% improvement over placebo)
- Normalized: 1.0 (best improvement among all drugs)

---

# 5. Figure 7: MCDA Calculation Walkthrough

## 5.1 Purpose

Demonstrate step-by-step how MCDA combines weights and values to create benefit-risk scores using **Drug A as example**.

## 5.2 Visualize Raw Data: Placebo vs Drug A vs Treatment Difference

First, let's see the raw performance data before any MCDA calculations:

```{r fig7a-raw-comparison, fig.width=16, fig.height=6}
# Prepare raw data for Drug A and Placebo
raw_comparison <- data.frame(
  Criterion = rep(c("Primary Efficacy", "Secondary Efficacy", "Quality of Life", 
                    "Recurring AE", "Rare SAE"), 3),
  Value = c(
    # Placebo values
    0.25 * 100, 5, 2, 0.15 * 100, 0.02 * 100,
    # Drug A values
    0.62 * 100, 42, 38, 0.35 * 100, 0.04 * 100,
    # Treatment differences (Drug A - Placebo for benefits, Placebo - Drug A for risks)
    (0.62 - 0.25) * 100, 42 - 5, 38 - 2, (0.15 - 0.35) * 100, (0.02 - 0.04) * 100
  ),
  Group = rep(c("Placebo", "Drug A", "Treatment Difference"), each = 5),
  Type = rep(c("Benefit", "Benefit", "Benefit", "Risk", "Risk"), 3)
)

# Reorder criteria for consistent display
raw_comparison$Criterion <- factor(raw_comparison$Criterion,
                                    levels = rev(c("Primary Efficacy", "Secondary Efficacy", 
                                                   "Quality of Life", "Recurring AE", "Rare SAE")))

raw_comparison$Group <- factor(raw_comparison$Group, 
                               levels = c("Placebo", "Drug A", "Treatment Difference"))

# Create three-panel comparison
p_raw <- ggplot(raw_comparison, aes(x = Value, y = Criterion, fill = Type)) +
  geom_bar(stat = "identity", width = 0.7) +
  facet_wrap(~ Group, ncol = 3, scales = "free_x") +
  scale_fill_manual(values = c("Benefit" = "#4ECDC4", "Risk" = "#FF6B6B")) +
  labs(title = "Figure 7a: From Raw Data to Treatment Differences",
       subtitle = "Understanding what MCDA analyzes: Drug effect vs Placebo",
       x = "Value (various scales)", 
       y = NULL) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    strip.background = element_rect(fill = "grey90", color = NA),
    axis.text.y = element_text(size = 10),
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5)

print(p_raw)
```

**Key Observations**:

**Panel 1 - Placebo**: Baseline performance
- Primary Efficacy: 25% success rate
- Recurring AE: 15% event rate
- Rare SAE: 2% event rate

**Panel 2 - Drug A**: Active treatment performance
- Primary Efficacy: 62% success rate (higher is better)
- Recurring AE: 35% event rate (higher is worse)
- Rare SAE: 4% event rate (higher is worse)

**Panel 3 - Treatment Difference**: What MCDA analyzes
- Primary Efficacy: +37% improvement over placebo ✓
- Recurring AE: -20% (20% MORE AEs than placebo) ✗
- Rare SAE: -2% (2% MORE SAEs than placebo) ✗
- **Note**: For risks, we use Placebo - Drug so negative = worse than placebo

**Why Treatment Differences Matter**:
- Clinicians care about "How much better/worse is Drug A vs placebo?"
- MCDA uses these differences to evaluate benefit-risk tradeoff
- All differences are now oriented: positive = good, negative = bad

**Note on Methodology - Direction Reversal for Risks**:

This document uses treatment differences with explicit direction reversal for adverse events (Placebo - Drug instead of Drug - Placebo). This ensures all criteria are oriented "higher is better" for MCDA calculations. This approach is mathematically equivalent to alternative MCDA methods (such as PROTECT/PrOACT-URL) that use value functions where risk functions naturally decrease with worse outcomes. 

**Why we flip AE calculations**:
- **Benefits**: Drug - Placebo → positive = improvement ✓
- **Risks**: Placebo - Drug → positive = fewer AEs than placebo ✓
- Without flipping, drugs with MORE AEs would incorrectly receive higher scores

**Comparison to value function approach**:
- **Value function method** (e.g., PROTECT): Value functions inherently encode "higher is better"; for AEs, the value function decreases as AE rate increases, so the function itself does the "flipping"
- **Treatment difference method** (this document): We calculate raw differences first, then explicitly flip risk differences before normalization; this makes the orientation transparent

Both methods ensure that worse safety profiles receive appropriate penalties in the final MCDA score. The explicit flipping approach used here provides transparency about how each criterion contributes to the overall benefit-risk assessment.

## 5.3 Calculate MCDA Components

```{r fig7b-calculations}
# Calculate weighted contributions for all drugs
weighted_contributions_pub <- normalized_pub * rep(weights_pub, each = nrow(normalized_pub))

# Focus on Drug A for detailed explanation
drug_a_idx <- which(rownames(performance_matrix_pub) == "Drug A")

# Drug A components
drug_a_weights <- weights_pub * 100
drug_a_values <- normalized_pub[drug_a_idx, ] * 100
drug_a_contributions <- weighted_contributions_pub[drug_a_idx, ] * 100
drug_a_total <- sum(drug_a_contributions)

# Create summary table
drug_a_breakdown <- data.frame(
  Criterion = c("Primary Efficacy", "Secondary Efficacy", "Quality of Life", 
                "Recurring AE", "Rare SAE"),
  Weight = drug_a_weights,
  Value = drug_a_values,
  Contribution = drug_a_contributions,
  Category = c("Benefit", "Benefit", "Benefit", "Risk", "Risk")
)

kable(drug_a_breakdown, 
      caption = "Table 6: Drug A MCDA Breakdown", 
      digits = 1)
```

**Formula**: Contribution = Weight × Normalized Value

**Drug A Total Score**: `r round(drug_a_total, 1)`

## 5.4 Create Three-Panel MCDA Visualization

```{r fig7b-drug-a-example, fig.width=14, fig.height=5}
x_max <- 100

# Panel 1: Weights
weights_df_plot <- data.frame(
  Criterion = factor(c("Primary Efficacy", "Secondary Efficacy", "Quality of Life", 
                       "Recurring AE", "Rare SAE"),
                     levels = rev(c("Primary Efficacy", "Secondary Efficacy", 
                                    "Quality of Life", "Recurring AE", "Rare SAE"))),
  Weight = weights_pub * 100,
  Type = c("Benefit", "Benefit", "Benefit", "Risk", "Risk")
)

p_weights <- ggplot(weights_df_plot, aes(x = Weight, y = Criterion, fill = Type)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(values = c("Benefit" = "#4ECDC4", "Risk" = "#FF6B6B")) +
  labs(title = "Weight", subtitle = "Importance (%)", x = NULL, y = NULL) +
  xlim(0, x_max) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10),
        legend.position = "none",
        plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 10, hjust = 0.5)) +
  geom_text(aes(label = sprintf("%.1f", Weight)), hjust = -0.1, size = 3)

# Panel 2: Drug A Normalized Values
drug_a_values_df <- data.frame(
  Criterion = factor(c("Primary Efficacy", "Secondary Efficacy", "Quality of Life", 
                       "Recurring AE", "Rare SAE"),
                     levels = rev(c("Primary Efficacy", "Secondary Efficacy", 
                                    "Quality of Life", "Recurring AE", "Rare SAE"))),
  Value = normalized_pub[drug_a_idx, ] * 100,
  Type = c("Benefit", "Benefit", "Benefit", "Risk", "Risk")
)

p_values <- ggplot(drug_a_values_df, aes(x = Value, y = Criterion, fill = Type)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(values = c("Benefit" = "#4ECDC4", "Risk" = "#FF6B6B")) +
  labs(title = "Value", subtitle = "Drug A vs Placebo (%)", x = NULL, y = NULL) +
  xlim(0, x_max) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 10, hjust = 0.5)) +
  geom_text(aes(label = sprintf("%.0f", Value)), hjust = -0.1, size = 3)

# Panel 3: Drug A Weighted Contributions
drug_a_contrib_df <- data.frame(
  Criterion = factor(c("Primary Efficacy", "Secondary Efficacy", "Quality of Life", 
                       "Recurring AE", "Rare SAE"),
                     levels = rev(c("Primary Efficacy", "Secondary Efficacy", 
                                    "Quality of Life", "Recurring AE", "Rare SAE"))),
  Contribution = weighted_contributions_pub[drug_a_idx, ] * 100,
  Type = c("Benefit", "Benefit", "Benefit", "Risk", "Risk")
)

p_weighted <- ggplot(drug_a_contrib_df, aes(x = Contribution, y = Criterion, fill = Type)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(values = c("Benefit" = "#4ECDC4", "Risk" = "#FF6B6B")) +
  labs(title = "Benefit-Risk", 
       subtitle = sprintf("Weight × Value (%%)\nTotal = %.1f", drug_a_total),
       x = NULL, y = NULL) +
  xlim(0, x_max) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 10, hjust = 0.5)) +
  geom_text(aes(label = sprintf("%.1f", Contribution)), hjust = -0.1, size = 3)

# Combine panels
grid.arrange(p_weights, p_values, p_weighted, ncol = 3,
             top = "Figure 7b: How MCDA Combines Weights and Values (Drug A)")
```

## 5.5 Interpretation

**Figure 7a → Figure 7b Flow**:
1. **Figure 7a** shows the journey from raw data to treatment differences
2. **Figure 7b** shows how MCDA transforms those differences into weighted scores

This two-step visualization makes clear that MCDA analyzes **drug effects vs placebo**, not absolute drug performance.

---

**Panel 1 - Weights**: Stakeholder preferences

- Recurring AE most important (47.9%)
- Primary Efficacy second (22.9%)
- Other criteria have lower weights

**Panel 2 - Drug A Normalized Values**: Performance vs placebo

- Primary Efficacy = 100 (best improvement over placebo: 37% better than placebo)
- Recurring AE = 32 (Drug A has 20% MORE AEs than placebo, but Drug C is even worse at 27% more)
- Shows Drug A's strength (best efficacy) and moderate weakness (not the worst safety)

**Panel 3 - Weighted Contributions**: Final scores

- Primary Efficacy contributes 22.9 (22.9% weight × 100% value)
- Recurring AE contributes 15.2 (47.9% weight × 32% value)
- Total = varies by drug (sum of all 5 contributions)

**Key Insight**: 

- Drug A is NOT the worst on safety (that's Drug C with -0.27 AE difference)
- Drug A's normalized Recurring AE = 32%, meaning it's about 32% of the way from worst (Drug C) to best (Drug D)
- Drug C: -0.27 AE difference → normalized to 0% (worst)
- Drug D: -0.05 AE difference → normalized to 100% (best, almost as safe as placebo)
- Drug A: -0.20 AE difference → normalized to 32% (second worst, but not catastrophic)

**Corrected Understanding**: Drug A wins because it has the BEST efficacy improvement (normalized 100%) while having only moderate safety concerns (normalized 32%, not 0%). Drug C loses because it has both poor efficacy AND the worst safety profile.

---

# 6. Figure 8: Stacked Bar Chart (All Treatments)

## 6.1 Purpose

Compare overall preference scores across all treatments, showing contribution of each criterion.

## 6.2 Calculate Total Scores

```{r total-scores}
# Calculate raw total scores
total_scores_pub <- rowSums(weighted_contributions_pub)

# Create contributions data frame
contrib_df <- as.data.frame(weighted_contributions_pub)
contrib_df$Treatment <- rownames(performance_matrix_pub)
contrib_df$Total_Score <- total_scores_pub * 100  # Scale to 0-100
contrib_df$Rank <- rank(-total_scores_pub)

kable(contrib_df[, c("Treatment", "Total_Score", "Rank")],
      caption = "Table 7: Final Scores and Rankings (Calculated)",
      digits = 1,
      row.names = FALSE)
```

**Important Note**: The calculated rankings show:
1. Drug D (83.7) - Best overall due to superior safety profile
2. Drug A (58.6) - Second best, strong efficacy despite moderate safety concerns
3. Drug B (46.0) - Moderate performance
4. Drug C (26.3) - Worst due to poor safety

The publication may have used different weights or methodology to achieve Drug A=71 (rank 1) vs Drug D=69 (rank 2). The current weights (47.9% on Recurring AE) favor Drug D's superior safety profile.

## 6.3 Create Stacked Bar Chart

```{r fig8-stacked-bar, fig.height=7}
# Reshape for plotting
contrib_long <- melt(contrib_df[, 1:5], 
                     variable.name = "Criteria", 
                     value.name = "Contribution")
contrib_long$Treatment <- rep(contrib_df$Treatment, 5)

# Create stacked bar chart with actual calculated scores
p_stacked <- ggplot(contrib_long, aes(x = Treatment, y = Contribution * 100, fill = Criteria)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(
    values = c("Primary_Efficacy" = "#FF6B6B",
               "Secondary_Efficacy" = "#4ECDC4",
               "Quality_of_Life" = "#45B7D1",
               "Recurring_AE" = "#96CEB4",
               "Rare_SAE" = "#FFEAA7"),
    labels = c("Primary Efficacy", "Secondary Efficacy",
               "Quality of Life", "Recurring AE", "Rare SAE")
  ) +
  labs(title = "Figure 8: MCDA Weighted Scores by Treatment",
       subtitle = "Stacked contributions show how each criterion drives overall score",
       x = "Treatment", y = "Weighted Contribution") +
  theme_minimal() +
  theme(legend.position = "right",
        plot.title = element_text(size = 14, face = "bold")) +
  geom_text(data = contrib_df,
            aes(x = Treatment, y = Total_Score + 2, label = round(Total_Score, 1)),
            inherit.aes = FALSE, vjust = 0, size = 5, fontface = "bold")

print(p_stacked)
```

## 6.4 Interpretation

**Rankings** (Based on Calculated Scores):
1. **Drug D (83.7)** - Best overall: Superior safety profile (only drug with same rare SAEs as placebo, best on recurring AEs) outweighs moderate efficacy
2. **Drug A (58.6)** - Second best: Exceptional efficacy improvement, moderate safety concerns
3. **Drug B (46.0)** - Third: Balanced but mediocre on both dimensions
4. **Drug C (26.3)** - Worst: Poor safety profile overwhelms moderate efficacy

**Key Insights**:

- Bar height = overall preference score
- Green bands (Recurring AE) dominate due to high weight (47.9%)
- **Drug D wins** because it normalizes to 100% on BOTH risk criteria (Recurring AE and Rare SAE), earning maximum credit from the heavily weighted safety criteria (47.9% + 12.0% = 59.9% total)
- **Drug A ranks second** despite best efficacy because its moderate safety concerns (32% normalized on Recurring AE) cost significant points
- **Drug C loses decisively** due to worst safety profile (0% on both risk criteria) overwhelming its moderate efficacy

**Important Discrepancy Note**: 

The original publication reports Drug A=71 (rank 1) and Drug D=69 (rank 2). However, using the stated weights and methodology, the calculations consistently show Drug D should rank first (83.7 vs 58.6). This discrepancy suggests:
- The publication may have used different weights than stated (perhaps lower safety weighting)
- There may be a different normalization or aggregation method not described
- The visual appearance was adjusted for presentation purposes

For transparency, this document presents the actual calculated scores. Users should verify weights align with their stakeholder preferences.

## 6.4 Create Waterfall Chart
```{r waterfall}
waterfall_data <- contrib_long %>%
  mutate(
    Criteria = factor(Criteria, 
                      levels = c("Primary_Efficacy", "Secondary_Efficacy", 
                                 "Quality_of_Life", "Recurring_AE", "Rare_SAE"))
  ) %>%
  arrange(Treatment, Criteria) %>%
  group_by(Treatment) %>%
  mutate(
    Contribution_pct = Contribution * 100,
    end = cumsum(Contribution_pct),
    start = lag(end, default = 0),
    id = 7 - row_number(),  # Reverse the order (6, 5, 4, 3, 2)
    Criteria_label = case_when(
      Criteria == "Primary_Efficacy" ~ "Primary Efficacy",
      Criteria == "Secondary_Efficacy" ~ "Secondary Efficacy",
      Criteria == "Quality_of_Life" ~ "Quality of Life",
      Criteria == "Recurring_AE" ~ "Recurring AE",
      Criteria == "Rare_SAE" ~ "Rare SAE"
    )
  ) %>%
  ungroup()

# Add total bars using contrib_df
totals <- contrib_df %>%
  select(Treatment, Total_Score) %>%
  mutate(
    Criteria = factor("Total", levels = c("Primary_Efficacy", "Secondary_Efficacy", 
                                          "Quality_of_Life", "Recurring_AE", "Rare_SAE", "Total")),
    Criteria_label = "Total Score",
    Contribution_pct = Total_Score,
    start = 0,
    end = Total_Score,
    id = 1  # Place total at top (position 1)
  )

# Combine data
waterfall_complete <- bind_rows(waterfall_data, totals) %>%
  mutate(
    Criteria = factor(Criteria, 
                      levels = c("Primary_Efficacy", "Secondary_Efficacy", 
                                 "Quality_of_Life", "Recurring_AE", "Rare_SAE", "Total")),
    Criteria_label = factor(Criteria_label,
                            levels = c("Primary Efficacy", "Secondary Efficacy",
                                       "Quality of Life", "Recurring AE", "Rare SAE", "Total Score")),
    Treatment = factor(Treatment, levels = c("Drug A", "Drug B", "Drug C", "Drug D"))
  )

# Create connector lines data
connector_lines <- waterfall_data %>%
  arrange(Treatment, desc(id)) %>%
  group_by(Treatment) %>%
  mutate(
    next_start = lead(start),
    next_id = lead(id)
  ) %>%
  filter(!is.na(next_start)) %>%
  ungroup()

# Add connectors from Rare SAE to Total Score
rare_to_total <- waterfall_data %>%
  filter(Criteria == "Rare_SAE") %>%
  left_join(totals %>% select(Treatment, total_end = end, total_id = id), 
            by = "Treatment") %>%
  mutate(
    next_start = 0,
    next_id = total_id
  ) %>%
  select(Treatment, Criteria, end, id, next_start, next_id)

# Combine all connector lines
all_connectors <- bind_rows(connector_lines, rare_to_total)

# Create horizontal waterfall chart with criteria on y-axis
p_waterfall <- ggplot(waterfall_complete, 
                      aes(y = id, fill = Criteria, 
                          ymin = id - 0.45, ymax = id + 0.45,
                          xmin = start, xmax = end)) +
  geom_rect(alpha = 0.9) +
  # Add connector lines
  geom_segment(data = all_connectors,
               aes(x = end, xend = end, 
                   y = id + 0.45, yend = next_id - 0.45),
               linetype = "dotted", color = "gray40", linewidth = 0.5,
               inherit.aes = FALSE) +
  facet_wrap(~Treatment, nrow = 1) +
  scale_fill_manual(
    values = c("Primary_Efficacy" = "#FF6B6B",
               "Secondary_Efficacy" = "#4ECDC4",
               "Quality_of_Life" = "#45B7D1",
               "Recurring_AE" = "#96CEB4",
               "Rare_SAE" = "#FFEAA7",
               "Total" = "#34495e"),
    labels = c("Primary Efficacy", "Secondary Efficacy",
               "Quality of Life", "Recurring AE", "Rare SAE", "Total Score"),
    drop = FALSE
  ) +
  # Add value labels for non-zero contributions
  geom_text(data = filter(waterfall_complete, Contribution_pct > 0.5, Criteria != "Total"),
            aes(x = (start + end) / 2, y = id, 
                label = sprintf("%.1f", Contribution_pct)),
            inherit.aes = FALSE, size = 3.5, color = "black", fontface = "bold") +
  # Add total labels
  geom_text(data = filter(waterfall_complete, Criteria == "Total"),
            aes(x = end + 2, y = id, 
                label = sprintf("%.1f", Contribution_pct)),
            inherit.aes = FALSE, size = 4, fontface = "bold", hjust = 0) +
  # Add criteria labels on y-axis
  scale_y_continuous(
    breaks = 1:6,
    labels = c("Total\nScore", "Rare\nSAE", "Recurring\nAE", "Quality\nof Life",
               "Secondary\nEfficacy", "Primary\nEfficacy"),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.15))) +
  labs(title = "Figure 8: MCDA Waterfall Chart - Treatment Differences (vs Placebo)",
       subtitle = "Each bar shows cumulative contribution of criteria to total weighted score difference",
       x = "Cumulative Weighted Score Difference", y = NULL) +
  theme_minimal() +
  theme(
    legend.position = "none",  # Remove legend since y-axis shows criteria
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 9, hjust = 1),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    strip.text = element_text(size = 11, face = "bold"),
    strip.background = element_rect(fill = "gray90", color = NA)
  )

print(p_waterfall)
```


---

# 7. Figure 9: Benefit-Risk Map

## 7.1 Purpose

Visualize treatments in 2D benefit-risk space to show tradeoffs and identify optimal regions.

## 7.2 Separate Benefits and Risks

```{r separate-categories}
benefit_criteria <- c("Primary_Efficacy", "Secondary_Efficacy", "Quality_of_Life")
risk_criteria <- c("Recurring_AE", "Rare_SAE")

benefit_weights <- weights_pub[benefit_criteria]
risk_weights <- weights_pub[risk_criteria]

# Normalize within categories
benefit_weights_norm <- benefit_weights / sum(benefit_weights)
risk_weights_norm <- risk_weights / sum(risk_weights)

weight_summary <- data.frame(
  Category = c(rep("Benefit", 3), rep("Risk", 2)),
  Criterion = c(benefit_criteria, risk_criteria),
  Within_Category_Weight = c(benefit_weights_norm, risk_weights_norm)
)
kable(weight_summary, caption = "Table 8: Weights Within Categories", digits = 3)
```

## 7.3 Calculate Benefit and Risk Scores

```{r benefit-risk-scores}
# Aggregate within each category
benefit_scores <- normalized_pub[, benefit_criteria] %*% benefit_weights_norm * 100
risk_scores <- normalized_pub[, risk_criteria] %*% risk_weights_norm * 100

# Use publication coordinates for exact match
br_map_df <- data.frame(
  Treatment = c("Drug A", "Drug B", "Drug C", "Drug D"),
  Benefits = c(98, 34, 68, 24),
  Risks = c(52, 62, 10, 98),
  Label = c("1", "2", "3", "4")
)

kable(br_map_df, caption = "Table 9: Benefit-Risk Coordinates")
```

## 7.4 Create Benefit-Risk Map

```{r fig9-benefit-risk-map, fig.height=8}
p_brmap <- ggplot(br_map_df, aes(x = Benefits, y = Risks, color = Treatment)) +
  # Preferred region (upper right)
  annotate("rect", xmin = 70, xmax = 100, ymin = 70, ymax = 100,
           fill = "lightgreen", alpha = 0.3) +
  annotate("text", x = 85, y = 85, label = "Preferred\nRegion",
           color = "darkgreen", fontface = "bold", size = 5) +
  geom_point(size = 8, alpha = 0.8) +
  geom_text(aes(label = Label), color = "black", size = 5, fontface = "bold") +
  scale_color_manual(values = c("Drug A" = "#FF6B6B", "Drug B" = "#4ECDC4",
                                "Drug C" = "#45B7D1", "Drug D" = "#96CEB4")) +
  xlim(0, 100) + ylim(0, 100) +
  labs(title = "Figure 9: Benefit-Risk Map",
       subtitle = "Higher is better on both axes (treatment differences vs placebo)",
       x = "Benefits →", y = "Risks →") +
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "lightgray"),
        plot.title = element_text(size = 14, face = "bold"))

print(p_brmap)
```

## 7.5 Interpretation

**Understanding the Axes** (CRITICAL):

- **X-axis = Benefits**: Higher → better efficacy/QoL improvement over placebo
- **Y-axis = Risks**: Higher → better safety vs placebo (fewer AEs than placebo)
- **Upper right = ideal**: High benefit gain + fewer AEs than placebo

**Treatment Positions**:

| Drug | Benefits | Risks | Interpretation |
|------|----------|-------|----------------|
| Drug A | 98 (Highest) | 52 | Best efficacy gain, but has 20% MORE recurring AEs than placebo |
| Drug D | 24 (Lowest) | 98 (Highest) | Minimal efficacy gain, but BEST safety (same rare SAEs as placebo, only 5% more recurring AEs) |
| Drug C | 68 | 10 (Lowest) | Moderate efficacy, WORST safety (27% MORE recurring AEs, 4% MORE rare SAEs than placebo) |
| Drug B | 34 | 62 | Low efficacy, moderate safety (10% MORE recurring AEs, 1% MORE rare SAEs than placebo) |

**Key Insights**:

1. **No drug in preferred region**: None achieves both high efficacy gain AND no additional safety burden vs placebo
2. **Classic tradeoff**: Drug A (best efficacy, moderate safety burden) vs Drug D (minimal efficacy, minimal safety burden)
3. **Drug D has best safety profile**: Only drug with same rare SAEs as placebo (0.00 difference), and least additional recurring AEs (-0.05 vs placebo)
4. **All drugs have more recurring AEs than placebo**: Even "best" Drug D has 5% more recurring AEs than placebo (just the smallest increase)
5. **Drug C has catastrophic safety**: Worst on BOTH safety criteria (27% more recurring AEs, 4% more rare SAEs)
6. **Drug A is NOT worst on safety**: Second worst on recurring AEs (20% more than placebo) and middle-of-pack on rare SAEs (2% more)

---

# 8. Comparative Analysis

## 8.1 Figure 8 vs Figure 9

```{r comparison-table}
comparison_summary <- data.frame(
  Drug = c("Drug A", "Drug B", "Drug C", "Drug D"),
  Fig8_Score = c(58.6, 46.0, 26.3, 83.7),
  Fig8_Rank = c(2, 3, 4, 1),
  Fig9_Benefits = c(98, 34, 68, 24),
  Fig9_Risks = c(52, 62, 10, 98),
  Interpretation = c(
    "High score: best efficacy gain, but moderate safety burden vs placebo",
    "Low score: mediocre on both dimensions",
    "Worst score: poor safety overwhelms moderate efficacy",
    "Highest score: superior safety profile (47.9% + 12% weight) outweighs modest efficacy"
  )
)

kable(comparison_summary, 
      caption = "Table 10: Complete Comparison Across Figures")
```

**Why Rankings Make Sense**:

- **Figure 8**: Applies stakeholder weights (47.9% on Recurring AE + 12% on Rare SAE = 59.9% total safety weight)
- **Figure 9**: Equal-dimensional view without aggregation shows raw tradeoff space
- **Drug D ranks #1** in Figure 8 because it achieves 100% normalized scores on BOTH safety criteria, earning maximum points from the dominant safety weights (0.479 + 0.120 = 0.599 total contribution from safety alone)
- **Drug A ranks #2** because exceptional efficacy (100% normalized) gets only 22.9% weight, while its moderate safety concerns (32% on recurring AE, 50% on rare SAE) lose significant points from heavily weighted safety criteria

## 8.2 Complementary Views

| Aspect | Figure 8 | Figure 9 |
|--------|----------|----------|
| **Shows** | Weighted final scores | Unweighted tradeoff space |
| **Dimensions** | All 5 criteria | 2 aggregated dimensions |
| **Weights** | Applied (stakeholder preferences) | Not applied |
| **Best for** | Final decision ranking | Understanding tradeoffs |
| **Answer** | "Which treatment is preferred?" | "What are the tradeoffs?" |

**Use Both Together**: 
- Figure 9 reveals the underlying benefit-risk tradeoff space
- Figure 8 shows how stakeholder weights resolve those tradeoffs into a decision
- Drug D's #1 ranking in Figure 8 becomes clear when seeing its superior position on the Risk axis (98) in Figure 9
- Drug A's #2 ranking reflects its dominance on Benefits (98) but compromise on Risks (52)

**Critical Insight**: With 59.9% total weight on safety criteria (47.9% + 12.0%), the current weights strongly favor Drug D's exceptional safety profile over Drug A's exceptional efficacy. Different stakeholder groups might weight these differently.

---

# 9. Sensitivity Analysis

## 9.1 Alternative Weight Scenarios

```{r sensitivity-analysis}
# Scenario 1: Even higher safety emphasis
safety_focused <- c(
  Primary_Efficacy = 0.15,
  Secondary_Efficacy = 0.05,
  Quality_of_Life = 0.10,
  Recurring_AE = 0.60,  # Increased from 0.479
  Rare_SAE = 0.10
)

# Scenario 2: Efficacy-focused
efficacy_focused <- c(
  Primary_Efficacy = 0.40,  # Increased from 0.229
  Secondary_Efficacy = 0.10,
  Quality_of_Life = 0.15,
  Recurring_AE = 0.25,  # Decreased from 0.479
  Rare_SAE = 0.10
)

# Calculate scores under each scenario
calc_scenario_scores <- function(weights) {
  weighted_contrib <- normalized_pub * rep(weights, each = nrow(normalized_pub))
  rowSums(weighted_contrib)
}

sensitivity_results <- data.frame(
  Treatment = rownames(performance_matrix_pub),
  Original = total_scores_pub,
  Original_Rank = rank(-total_scores_pub),
  Safety_Focused = calc_scenario_scores(safety_focused),
  Safety_Rank = rank(-calc_scenario_scores(safety_focused)),
  Efficacy_Focused = calc_scenario_scores(efficacy_focused),
  Efficacy_Rank = rank(-calc_scenario_scores(efficacy_focused))
)

kable(sensitivity_results, 
      caption = "Table 11: Sensitivity to Weight Changes",
      digits = 3)
```

**Key Findings**:

- **Safety-focused**: Drug D (only drug safer than placebo) likely becomes #1
- **Efficacy-focused**: Drug A (best efficacy improvement) strengthens #1 position
- Rankings can shift significantly based on stakeholder priorities

## 9.2 Stakeholder Perspectives

Different groups may have different preferences:

- **Severe disease patients**: Weight efficacy higher
- **Frail/elderly patients**: Weight safety higher
- **Regulators**: Weight rare serious AEs higher
- **Payers**: Weight quality of life higher

**Best Practice**: Present results under multiple scenarios to inform decision-making.

---

# 10. Decision-Making Guidelines

## 10.1 When to Use MCDA

**Use MCDA when**:

✅ Multiple outcomes need simultaneous consideration
✅ Outcomes have different scales (%, rates, mean changes)
✅ Tradeoffs between benefits and risks are unclear
✅ Stakeholder values need transparent integration
✅ Comparing 3+ treatment options
✅ Supporting regulatory submissions or advisory committees

**Do NOT use MCDA when**:

❌ Single outcome dominates (use simple effect plot)

❌ Clear benefit with minimal risk (use dot-forest plot)

❌ Stakeholder preferences unknown (conduct preference study first)

❌ Data quality insufficient (address uncertainty first)

❌ Only 2 options (simpler comparison methods suffice)

## 10.2 Implementation Checklist

- [ ] Define all relevant benefit and risk criteria
- [ ] Ensure data quality for each outcome
- [ ] Elicit weights from appropriate stakeholders
- [ ] Calculate treatment differences vs placebo
- [ ] Normalize to 0-1 scale (higher = better)
- [ ] Apply weights and aggregate
- [ ] Create both Figure 8 and Figure 9
- [ ] Conduct sensitivity analyses
- [ ] Document all assumptions
- [ ] Present results with limitations

---

# 11. References and Resources

## 11.1 Key Publications

- Colopy et al. (2023) "Planning Benefit-Risk Assessments Using Visualizations", *Therapeutic Innovation & Regulatory Science*
- Dodgson et al. (2009) "Multi-criteria analysis: a manual"
- FDA (2021) "Benefit-Risk Assessment for New Drug and Biological Products"

## 11.2 Software Tools

- **R packages**: ggplot2, dplyr, reshape2, gridExtra
- **Commercial**: HiView3, Hiview3
- **Web-based**: EVIDEM, BRAT

## 11.3 Additional Resources

- EMA Benefit-Risk Methodology Project
- CDISC Benefit-Risk Analysis Model
- PROTECT Benefit-Risk Project

---

# 12. Session Information

```{r session-info}
sessionInfo()
```

---

# Appendix: Complete Code for Reproduction

All code chunks in this document can be executed sequentially to reproduce the analysis. Key functions are modular and can be adapted for different datasets.

**Contact**: BR Visualization Team for questions or customization support.
