---
title: "MCDA Clinical Threshold-Based Normalization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCDA Clinical Threshold-Based Normalization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 14,
  fig.height = 6
)
```

```{r setup, eval=FALSE}
library(brpubVJCE)
```

## Overview

The `create_mcda_barplot_walkthrough()` function implements **clinical threshold-based normalization** for Multi-Criteria Decision Analysis (MCDA). This approach uses fixed clinical scales (global scales) rather than treatment-relative normalization (local scales), as recommended by FDA/EMA best practices and the PROTECT framework.

## The Normalization Approach

### Value Function

For each criterion, we apply a linear value function based on clinical thresholds:

**For "increasing" direction (higher is better):**
```r
v(x) = 100 * (x - min) / (max - min)
```

**For "decreasing" direction (lower is better):**
```r
v(x) = 100 * (max - x) / (max - min)
```

Where `min` and `max` are clinically meaningful thresholds, not observed treatment values.

### Key Steps

1. **Normalize actual values separately**
   - Apply value function to drug's actual values → Drug normalized (0-100)
   - Apply value function to placebo's actual values → Placebo normalized (0-100)

2. **Compute normalized differences**
   - Normalized difference = Drug normalized - Placebo normalized
   - Positive = Drug performs better than Placebo
   - Negative = Drug performs worse than Placebo

3. **Apply weights and aggregate**
   - Weighted contribution = Normalized difference × Weight
   - Total benefit-risk score = Sum of weighted contributions

## Defining Clinical Scales

Each criterion requires three parameters:

```r
clinical_scales <- list(
  `Criterion Name` = list(
    min = 0,                  # Lower threshold
    max = 100,                # Upper threshold
    direction = "increasing"  # "increasing" or "decreasing"
  )
)
```

**Direction:**
- `"increasing"`: Higher raw values are better (e.g., efficacy)
- `"decreasing"`: Lower raw values are better (e.g., adverse events)

Thresholds should be based on clinical guidelines, Minimum Clinically Important Difference (MCID), or regulatory precedents.

## Example

### Prepare Data

```{r example-data, eval=FALSE}
mcda_data <- prepare_mcda_data(effects_table)
```

### Define Clinical Scales

```{r clinical-scales, eval=FALSE}
clinical_scales <- list(
  `Benefit 1` = list(min = 0, max = 1, direction = "increasing"),
  `Benefit 2` = list(min = 0, max = 100, direction = "decreasing"),
  `Benefit 3` = list(min = 0, max = 100, direction = "increasing"),
  `Risk 1` = list(min = 0, max = 0.5, direction = "decreasing"),
  `Risk 2` = list(min = 0, max = 0.15, direction = "decreasing")
)
```

### Define Weights

```{r weights, eval=FALSE}
weights <- c(
  `Benefit 1` = 0.30,
  `Benefit 2` = 0.20,
  `Benefit 3` = 0.10,
  `Risk 1` = 0.30,
  `Risk 2` = 0.10
)
```

### Create Visualization

```{r visualization, eval=FALSE}
barplot_walkthrough <- create_mcda_barplot_walkthrough(
  data = mcda_data,
  benefit_criteria = c("Benefit 1", "Benefit 2", "Benefit 3"),
  risk_criteria = c("Risk 1", "Risk 2"),
  comparison_drug = "Drug A",
  weights = weights,
  clinical_scales = clinical_scales
)
```

## Interpreting the Output

The plot shows 4 panels representing the MCDA calculation steps:

**Panel 1: Treatment Difference (Raw)**
- Shows: Drug - Placebo in original units
- Note: Sign doesn't indicate better/worse

**Panel 2: Normalized Difference**
- Shows: (Drug normalized - Placebo normalized) on 0-100 scale
- Positive = Drug performs better
- Negative = Drug performs worse

**Panel 3: Weight**
- Shows: Relative importance (must sum to 100%)

**Panel 4: Benefit-Risk**
- Shows: Normalized Difference × Weight
- Total score = Sum of contributions
- Positive total = Drug better overall

## Understanding Negative Normalized Values

Example: Drug has 19% adverse events, Placebo has 3% adverse events

```r
# Clinical scale: min=0%, max=50%, direction="decreasing"

# Step 1: Normalize separately
Drug_normalized = 100 * (50-19) / 50 = 62
Placebo_normalized = 100 * (50-3) / 50 = 94

# Step 2: Compute difference
Normalized_difference = 62 - 94 = -32  # Drug performs worse
```

The -32 indicates Drug has worse safety than Placebo (higher AE rate)

## Key Points

**Setting Thresholds:**
- Base on clinical guidelines, MCID, or regulatory precedents
- Do NOT use observed min/max from your data
- Document rationale

**Validation:**
- Check face validity (does it make clinical sense?)
- Perform sensitivity analysis on weights
- Ensure stakeholder acceptance

## References

- Mussen F, et al. (2007). *Pharmacoepidemiol Drug Saf*, 16 Suppl 1:S2-15
- PROTECT Benefit-Risk Framework (2012)
- Thokala P, et al. (2016). *Value in Health*, 19(1):1-13
- FDA Benefit-Risk Assessment Framework (2013)
- EMA Benefit-Risk Methodology Project (2011)

For detailed implementation reference: `dev/references/MCDA_Value_Function_Implementation.md`
