---
title: "MCDA Value Function Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCDA Value Function Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
```

## Introduction

Value functions are mathematical transformations that convert raw clinical outcome values into a standardized scale (typically 0-100) representing "value" or "desirability" from a decision-making perspective. The choice of value function reflects stakeholder preferences about how outcomes are valued.

## Why Value Functions Matter

Consider an adverse event rate:
- **Linear**: Going from 10% to 20% AE rate has the same negative value as 20% to 30%
- **Exponential**: Going from 10% to 20% is concerning, but 20% to 30% is dramatically worse
- **Threshold**: Below 15% is acceptable (value=100), above 15% is unacceptable (value=0)

The choice affects benefit-risk conclusions and should reflect clinical reality and stakeholder preferences.

## Value Function Types

### 1. Linear Value Functions (Current Standard)

**Description**: Constant marginal value - each unit change has equal importance

**Increasing direction** (higher is better, e.g., efficacy):
$$v(x) = 100 \times \frac{x - \text{min}}{\text{max} - \text{min}}$$

**Decreasing direction** (lower is better, e.g., adverse events):
$$v(x) = 100 \times \frac{\text{max} - x}{\text{max} - \text{min}}$$

```{r linear-example}
# Define linear value functions
linear_increasing <- function(x, min_val, max_val) {
  100 * (x - min_val) / (max_val - min_val)
}

linear_decreasing <- function(x, min_val, max_val) {
  100 * (max_val - x) / (max_val - min_val)
}

# Example: Efficacy (0-100% response rate)
x_efficacy <- seq(0, 100, by = 1)
value_efficacy <- linear_increasing(x_efficacy, min_val = 0, max_val = 100)

# Example: Adverse events (0-50% AE rate)
x_ae <- seq(0, 50, by = 0.5)
value_ae <- linear_decreasing(x_ae, min_val = 0, max_val = 50)

# Create plots
p_efficacy <- ggplot(data.frame(x = x_efficacy, value = value_efficacy),
                     aes(x = x, y = value)) +
  geom_line(color = "#0571b0", linewidth = 1.2) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "gray50") +
  labs(
    title = "Linear Increasing: Efficacy",
    x = "Response Rate (%)",
    y = "Value (0-100)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))

p_ae <- ggplot(data.frame(x = x_ae, value = value_ae),
               aes(x = x, y = value)) +
  geom_line(color = "#ca0020", linewidth = 1.2) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "gray50") +
  labs(
    title = "Linear Decreasing: Adverse Events",
    x = "AE Rate (%)",
    y = "Value (0-100)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))

p_efficacy + p_ae
```

**When to use:**
- Default choice (FDA/EMA recommendation)
- No strong evidence of non-linear preferences
- Transparency and simplicity are priorities
- Most clinical outcomes

**Advantages:**
- Simple and interpretable
- Transparent calculations
- Regulatory acceptance
- Conservative (neutral) assumptions

**Disadvantages:**
- May not reflect true stakeholder preferences
- Assumes equal marginal value across range

---

### 2. Piecewise Linear Value Functions

**Description**: Different slopes in different regions, connected at breakpoints

```{r piecewise-example}
# Piecewise linear function with 3 segments
piecewise_linear <- function(x, breakpoints, slopes, intercepts) {
  value <- numeric(length(x))
  for (i in seq_along(x)) {
    if (x[i] < breakpoints[1]) {
      value[i] <- slopes[1] * x[i] + intercepts[1]
    } else if (x[i] < breakpoints[2]) {
      value[i] <- slopes[2] * (x[i] - breakpoints[1]) + intercepts[2]
    } else {
      value[i] <- slopes[3] * (x[i] - breakpoints[2]) + intercepts[3]
    }
  }
  # Constrain to 0-100
  pmax(0, pmin(100, value))
}

# Example: AE severity with different concern levels
# Mild (0-10%): slope = -2
# Moderate (10-25%): slope = -3
# Severe (>25%): slope = -5

x_ae_pw <- seq(0, 40, by = 0.5)

# Calculate intercepts to ensure continuity
# At x=0: value=100
# At x=10: value=100-2*10=80
# At x=25: value=80-3*(25-10)=35

breakpoints <- c(10, 25)
slopes <- c(-2, -3, -5)
intercepts <- c(100, 80, 35)

value_ae_pw <- piecewise_linear(x_ae_pw, breakpoints, slopes, intercepts)

# Compare with linear
value_ae_linear <- linear_decreasing(x_ae_pw, min_val = 0, max_val = 50)

comparison_df <- data.frame(
  x = rep(x_ae_pw, 2),
  value = c(value_ae_pw, value_ae_linear),
  type = rep(c("Piecewise", "Linear"), each = length(x_ae_pw))
)

ggplot(comparison_df, aes(x = x, y = value, color = type, linetype = type)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = breakpoints, linetype = "dotted", color = "gray40") +
  annotate("text", x = 5, y = 95, label = "Mild", size = 3.5, color = "gray30") +
  annotate("text", x = 17.5, y = 95, label = "Moderate", size = 3.5, color = "gray30") +
  annotate("text", x = 32, y = 95, label = "Severe", size = 3.5, color = "gray30") +
  scale_color_manual(values = c("Piecewise" = "#ca0020", "Linear" = "gray50")) +
  scale_linetype_manual(values = c("Piecewise" = "solid", "Linear" = "dashed")) +
  labs(
    title = "Piecewise Linear: AE Severity Regions",
    x = "Adverse Event Rate (%)",
    y = "Value (0-100)",
    color = "Function",
    linetype = "Function"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

**When to use:**
- Clinical guidelines define severity categories (mild/moderate/severe)
- Regulatory thresholds exist (acceptable/concerning/unacceptable)
- Stakeholders value ranges differently
- MCID (Minimally Clinically Important Difference) creates natural breakpoints

**Example applications:**
- QoL scales with clinical interpretation bands
- Lab values with normal/borderline/abnormal ranges
- Dose levels with safety thresholds

**Advantages:**
- Captures clinical interpretation thresholds
- More flexible than linear
- Still relatively transparent
- Can match stakeholder preferences better

**Disadvantages:**
- More parameters to justify
- Arbitrary breakpoint placement needs rationale
- More complex than linear

---

### 3. Exponential/Power Value Functions

**Description**: Curved relationship reflecting risk attitudes

**Risk-averse (diminishing returns)**, b > 1:
$$v(x) = 100 \times \left(\frac{x - \text{min}}{\text{max} - \text{min}}\right)^b$$

**Risk-seeking (increasing returns)**, 0 < b < 1:
$$v(x) = 100 \times \left(\frac{x - \text{min}}{\text{max} - \text{min}}\right)^b$$

```{r exponential-example}
# Power/exponential functions
power_function <- function(x, min_val, max_val, power) {
  normalized <- (x - min_val) / (max_val - min_val)
  100 * normalized^power
}

x_vals <- seq(0, 100, by = 1)

# Different power values
powers <- c(0.5, 1, 2, 3)
power_data <- data.frame()

for (p in powers) {
  values <- power_function(x_vals, min_val = 0, max_val = 100, power = p)
  power_data <- rbind(
    power_data,
    data.frame(
      x = x_vals,
      value = values,
      power = as.factor(sprintf("b = %.1f", p))
    )
  )
}

ggplot(power_data, aes(x = x, y = value, color = power)) +
  geom_line(linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  annotate("text", x = 75, y = 85, label = "b < 1: Risk-seeking\n(increasing returns)",
           size = 3.5, color = "gray30", hjust = 0) +
  annotate("text", x = 25, y = 20, label = "b > 1: Risk-averse\n(diminishing returns)",
           size = 3.5, color = "gray30", hjust = 0) +
  scale_color_manual(values = c(
    "b = 0.5" = "#2c7bb6",
    "b = 1.0" = "gray50",
    "b = 2.0" = "#fdae61",
    "b = 3.0" = "#d7191c"
  )) +
  labs(
    title = "Power Functions: Risk Attitudes",
    subtitle = "b = 1 is linear (neutral)",
    x = "Raw Outcome Value",
    y = "Value (0-100)",
    color = "Power"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

**When to use:**
- Stakeholder elicitation reveals non-linear preferences
- Diminishing returns clinically meaningful (first 50% efficacy more valuable)
- Severity increases exponentially (rare but catastrophic events)
- Utility theory supports risk attitudes

**Example applications:**
- Mortality (each life equally valuable - linear or power=1)
- Quality of life (diminishing returns at high levels)
- Rare serious AEs (exponential concern)

**Advantages:**
- Captures risk attitudes
- Single parameter (power) to adjust
- Smooth and differentiable

**Disadvantages:**
- Requires justification for non-linearity
- Power parameter selection subjective
- Less transparent than linear

---

### 4. Sigmoid (S-shaped) Value Functions

**Description**: Slow change at extremes, rapid change in middle

$$v(x) = \frac{100}{1 + \exp(-k \times (x - \text{midpoint}))}$$

```{r sigmoid-example}
# Sigmoid/logistic function
sigmoid_function <- function(x, min_val, max_val, midpoint, steepness) {
  # Normalize x to midpoint
  x_centered <- x - midpoint
  # Apply logistic function
  100 / (1 + exp(-steepness * x_centered))
}

x_vals <- seq(0, 100, by = 0.5)

# Different steepness values (k parameter)
steepness_vals <- c(0.05, 0.1, 0.2)
sigmoid_data <- data.frame()

for (k in steepness_vals) {
  values <- sigmoid_function(x_vals, min_val = 0, max_val = 100,
                             midpoint = 50, steepness = k)
  sigmoid_data <- rbind(
    sigmoid_data,
    data.frame(
      x = x_vals,
      value = values,
      steepness = as.factor(sprintf("k = %.2f", k))
    )
  )
}

# Add linear for comparison
sigmoid_data <- rbind(
  sigmoid_data,
  data.frame(
    x = x_vals,
    value = linear_increasing(x_vals, 0, 100),
    steepness = "Linear"
  )
)

ggplot(sigmoid_data, aes(x = x, y = value, color = steepness, linetype = steepness)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 50, linetype = "dotted", color = "gray40") +
  geom_hline(yintercept = 50, linetype = "dotted", color = "gray40") +
  annotate("text", x = 50, y = 5, label = "Midpoint", size = 3, color = "gray30") +
  scale_color_manual(values = c(
    "k = 0.05" = "#2c7bb6",
    "k = 0.10" = "#fdae61",
    "k = 0.20" = "#d7191c",
    "Linear" = "gray50"
  )) +
  scale_linetype_manual(values = c(
    "k = 0.05" = "solid",
    "k = 0.10" = "solid",
    "k = 0.20" = "solid",
    "Linear" = "dashed"
  )) +
  labs(
    title = "Sigmoid (S-shaped) Value Functions",
    subtitle = "Threshold effect centered at midpoint = 50",
    x = "Raw Outcome Value",
    y = "Value (0-100)",
    color = "Steepness",
    linetype = "Steepness"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

**When to use:**
- Threshold effects (outcome "tips" at clinical cutpoint)
- Very low/high values matter less than mid-range
- Psychological preferences show S-shaped patterns
- Binary clinical decisions with gray zone

**Example applications:**
- Biomarkers with clinical interpretation thresholds
- PROs with "clinically meaningful change" bands
- Dose-response with therapeutic window

**Advantages:**
- Captures threshold effects
- Smooth transition (no sharp jumps)
- Biologically/clinically realistic for many outcomes

**Disadvantages:**
- Two parameters (midpoint, steepness) to justify
- Can be complex to explain
- Requires clear rationale for threshold location

---

### 5. Step Functions (Discrete Categories)

**Description**: Discrete value levels for categorical outcomes

```{r step-example}
# Step function for severity categories
step_function <- function(x, thresholds, values) {
  result <- numeric(length(x))
  for (i in seq_along(x)) {
    # Find which category x[i] falls into
    category <- sum(x[i] >= thresholds) + 1
    result[i] <- values[category]
  }
  result
}

x_vals <- seq(0, 50, by = 0.5)

# Example: AE severity categories
# None (0-5%): value=100
# Mild (5-15%): value=75
# Moderate (15-30%): value=40
# Severe (>30%): value=0

thresholds <- c(5, 15, 30)
values <- c(100, 75, 40, 0)

value_step <- step_function(x_vals, thresholds, values)
value_linear <- linear_decreasing(x_vals, 0, 50)

comparison_step <- data.frame(
  x = rep(x_vals, 2),
  value = c(value_step, value_linear),
  type = rep(c("Step", "Linear"), each = length(x_vals))
)

ggplot(comparison_step, aes(x = x, y = value, color = type, linetype = type)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = thresholds, linetype = "dotted", color = "gray40") +
  annotate("text", x = 2.5, y = 105, label = "None", size = 3.5, color = "gray30") +
  annotate("text", x = 10, y = 105, label = "Mild", size = 3.5, color = "gray30") +
  annotate("text", x = 22.5, y = 105, label = "Moderate", size = 3.5, color = "gray30") +
  annotate("text", x = 40, y = 105, label = "Severe", size = 3.5, color = "gray30") +
  scale_color_manual(values = c("Step" = "#ca0020", "Linear" = "gray50")) +
  scale_linetype_manual(values = c("Step" = "solid", "Linear" = "dashed")) +
  labs(
    title = "Step Function: Discrete Severity Categories",
    x = "Adverse Event Rate (%)",
    y = "Value (0-100)",
    color = "Function",
    linetype = "Function"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

**When to use:**
- Outcomes are inherently categorical (CTCAE grades)
- Clinical practice uses discrete classifications
- No meaningful distinction within categories
- Regulatory guidance defines categories

**Example applications:**
- CTCAE toxicity grades (1-5)
- ECOG performance status (0-4)
- Disease severity classifications

**Advantages:**
- Matches clinical practice
- Simple interpretation
- Natural for categorical data

**Disadvantages:**
- Ignores variation within categories
- Sharp discontinuities may not reflect reality
- Can be sensitive to threshold placement

---

## Comparing Value Functions

Let's visualize all function types for the same outcome:

```{r comparison-all}
x_vals <- seq(0, 50, by = 0.5)

# Generate all function types
comparison_all <- data.frame(
  x = rep(x_vals, 5),
  value = c(
    linear_decreasing(x_vals, 0, 50),
    piecewise_linear(x_vals, c(10, 25), c(-2, -3, -5), c(100, 80, 35)),
    power_function(50 - x_vals, 0, 50, 2),  # Decreasing + risk-averse
    100 - sigmoid_function(x_vals, 0, 50, 25, 0.15),  # Inverted sigmoid
    step_function(x_vals, c(5, 15, 30), c(100, 75, 40, 0))
  ),
  type = rep(
    c("Linear", "Piecewise Linear", "Exponential (b=2)",
      "Sigmoid", "Step"),
    each = length(x_vals)
  )
)

comparison_all$type <- factor(
  comparison_all$type,
  levels = c("Linear", "Piecewise Linear", "Exponential (b=2)",
             "Sigmoid", "Step")
)

ggplot(comparison_all, aes(x = x, y = value, color = type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c(
    "Linear" = "gray50",
    "Piecewise Linear" = "#2c7bb6",
    "Exponential (b=2)" = "#fdae61",
    "Sigmoid" = "#d7191c",
    "Step" = "#542788"
  )) +
  labs(
    title = "Comparison of Value Function Types",
    subtitle = "Adverse event rate: Lower is better",
    x = "Adverse Event Rate (%)",
    y = "Value (0-100)",
    color = "Function Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

**Key observations:**
- **Linear**: Constant concern across all levels
- **Piecewise**: Different concern in different ranges
- **Exponential**: Increasing concern (risk-averse)
- **Sigmoid**: Threshold effect at ~25%
- **Step**: Categorical interpretation

---

## Regulatory and Practical Considerations

### FDA/EMA Guidance

**Default recommendation: Linear value functions**
- Transparent and conservative
- Neutral assumption (no risk attitude imposed)
- Easiest to justify and explain

**When non-linear may be acceptable:**
1. Strong clinical/theoretical rationale
2. Validated through stakeholder elicitation
3. Pre-specified in analysis plan
4. Sensitivity analysis shows robustness
5. Documented justification

### Selection Criteria

**Clinical validity:**
- Does the function match clinical understanding?
- Are there clinical thresholds or categories?
- Do clinicians think linearly or non-linearly about this outcome?

**Stakeholder preferences:**
- Elicit preferences using standard techniques
- Choice experiments
- Swing weighting
- Direct function specification

**Data requirements:**
- Linear: Only min/max thresholds
- Piecewise: Breakpoints and slopes
- Exponential: Power parameter
- Sigmoid: Midpoint and steepness
- Step: Category thresholds and values

**Interpretability:**
- Can you explain it to regulators?
- Can patients understand it?
- Does it make intuitive sense?

### Sensitivity Analysis

Always perform sensitivity analysis:

```{r sensitivity-example, fig.height=8}
# Sensitivity to power parameter
x_vals <- seq(0, 100, by = 1)
powers <- c(0.5, 1, 1.5, 2, 2.5, 3)

sensitivity_data <- data.frame()
for (p in powers) {
  values <- power_function(x_vals, 0, 100, p)
  sensitivity_data <- rbind(
    sensitivity_data,
    data.frame(
      x = x_vals,
      value = values,
      power = p,
      power_label = sprintf("b = %.1f", p)
    )
  )
}

# Calculate area under curve as a sensitivity metric
auc_values <- sensitivity_data %>%
  group_by(power_label) %>%
  summarise(auc = sum(value) / n(), .groups = "drop")

p1 <- ggplot(sensitivity_data, aes(x = x, y = value, color = as.factor(power))) +
  geom_line(linewidth = 1.2) +
  scale_color_brewer(palette = "RdYlBu", direction = -1) +
  labs(
    title = "Sensitivity to Power Parameter",
    x = "Raw Outcome Value",
    y = "Value (0-100)",
    color = "Power (b)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))

p2 <- ggplot(auc_values, aes(x = power_label, y = auc, fill = power_label)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f", auc)), vjust = -0.5, size = 3.5) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  labs(
    title = "Average Value (Area Under Curve)",
    x = "Power Parameter",
    y = "Mean Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )

p1 / p2
```

---

## Recommendations

### 1. Start with Linear
- Default choice unless clear rationale for non-linear
- Easiest to justify and defend
- Regulatory preference

### 2. Consider Non-linear When:
- Clinical thresholds exist (→ piecewise or step)
- Stakeholder preferences clearly non-linear (→ power or sigmoid)
- Strong theoretical rationale (→ appropriate function type)
- Validated through preference elicitation

### 3. Document Everything
- Rationale for function choice
- Parameter selection process
- Stakeholder input
- Sensitivity analyses
- Clinical interpretation

### 4. Validate Choices
- Face validity with clinicians
- Stakeholder review
- Sensitivity analysis
- Comparison with alternative functions

### 5. Be Conservative
- When in doubt, use linear
- Simpler is better for communication
- Consider audience (regulators, patients, clinicians)

---

## References

### Methodological

- Thokala P, et al. (2016). Multiple criteria decision analysis for health care decision making. *Value in Health*, 19(1):1-13.
- Keeney RL, Raiffa H. (1993). *Decisions with Multiple Objectives: Preferences and Value Trade-Offs*. Cambridge University Press.
- Dyer JS, Sarin RK. (1979). Measurable multiattribute value functions. *Operations Research*, 27(4):810-822.

### Regulatory

- FDA. (2013). Structured Approach to Benefit-Risk Assessment in Drug Regulatory Decision-Making.
- EMA. (2011). Benefit-Risk Methodology Project: Report of the BRMWP Task Force.
- PROTECT. (2012). Work Package 5: Benefit-Risk Integration and Representation.

### Applications

- Mussen F, et al. (2007). Structured benefit-risk assessment for medicinal products. *Pharmacoepidemiol Drug Saf*, 16(S1):S2-15.
- Mt-Isa S, et al. (2014). Balancing benefit and risk of medicines. *Clinical Pharmacology & Therapeutics*, 96(4):438-446.
- Levitan B, et al. (2018). Application of the BRAT framework to case studies. *Pharmacoepidemiol Drug Saf*, 27(11):1223-1234.

---

## Implementation Note

The current `brpubVJCE` package implements **linear value functions** by default, which is consistent with FDA/EMA guidance and provides a transparent, conservative approach to benefit-risk assessment.

If you need to implement alternative value functions for specific applications, the examples in this vignette provide the mathematical framework. Always ensure proper documentation and justification for non-linear choices.
